---
title: "Data exploration for Heart diseases and common habits"
author: 
  - name: "Jose Daniel Trujillo Sandoval"
    orcid: "0009-0008-2978-2604"
    email: "jotr4809@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
editor: visual
---

# Introduction

This project would focus on assessing habits related to heart diseases. The data set contains information of random patients, Systolic and Diastolic heart rates, Body mass index, height and weight, and common habits like drinking and smoking

# Methods

## Reading the Data

Loading Packages

```{r}
# Load Packages
library(tidyverse)
library(here) 
# Read Data 
Cardiac_raw_data <- read.csv(here::here("data/raw/cardio_raw_data.csv"))
```

## Data Exploration Approach

```{r}

# Viewing the first rows 
head(Cardiac_raw_data) #Looks like everything is good at the first 7 rows
glimpse(Cardiac_raw_data) # Age: int, gender: chr, height: int, weight: dbl, Sys: int, dias: int, Cholesterol: chr, Glucose: chr, smoke: chr, alcohol: chr, physical: chr, disease: chr, bmi: dbl

# Checking dimensions 
dim(Cardiac_raw_data) #68,675 rows, 13 colums 

# looking at ages 
Cardiac_raw_data |> 
  count(age) 

# looking at height
Cardiac_raw_data |> 
  count(height) 

# looking at weight
Cardiac_raw_data |> 
  count(weight) 

# Grouping by height and weithg to look at the mean ages
Cardiac_raw_data |> 
  group_by(height, weight) |>
  summarise(mean_age = mean(age))

```

## Initial Data Tidying

# Results

\[This will be the core of your analysis with specific requirements\]

# Conclusions

## Summary of Findings

## Questions and Next Steps
