---
title: "Data exploration for Heart diseases and common habits"
author: 
  - name: "Jose Daniel Trujillo Sandoval"
    orcid: "0009-0008-2978-2604"
    email: "jotr4809@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: 
  html:
    toc: true
    toc-depth: 4
abstract: 
  This project explores the relationship between various health indicators, lifestyle habits, and the prevalence of cardiovascular diseases (CVDs) using a public dataset obtained from a GitHub repository. The analysis focused on key variables including age, systolic and diastolic blood pressure, cholesterol levels, glucose, BMI, and habits such as smoking, alcohol consumption, and physical activity. Data cleaning and transformation were performed to ensure consistency, and categorical variables were properly labeled and converted for analysis. 
  The findings reveal that CVD prevalence increases significantly with age, especially in individuals over 56 years. People with high or normal cholesterol levels and elevated blood pressure tend to show a higher presence of CVDs, while those with extremely high cholesterol levels showed no clear difference. Furthermore, physical inactivity was strongly associated with the presence of cardiovascular disease, more so than smoking or alcohol intake. Additionally, individuals with CVD often presented with hypertension and BMI values indicating overweight or obesity. 
  Future research could focus on the impact of height and weight, the role of glucose and diabetes, and the most common unhealthy behaviors by gender to further understand risk factors related to cardiovascular health.
license: "CC BY"
citation: 
  container-title: My Blog
  url: https://www.example.com
bibliography: references.bib
csl: apa.csl
editor: visual
execute:
  warning: false
---

------------------------------------------------------------------------

# Introduction

This project focuses on assessing habits associated with heart disease. The dataset includes information from randomly selected patients, covering systolic and diastolic blood pressure, body mass index (BMI), height and weight, as well as common lifestyle habits such as alcohol consumption and smoking.

Throughout the project, we analyzed the prevalence of unhealthy habits and how these correlate with the presence of cardiovascular diseases (CVDs). The initial dataset was obtained from a public GitHub repository.

# Methods

## Reading the Data

Here we present all the libraries that are going to be used along the project, and also the location of the initial data set.

Loading Packages

```{r}
# Load Packages
library(tidyverse)
library(here) 
library(ggthemes)
library(gt)
# Read Data 
Cardiac_raw_data <- read.csv(here::here("data/raw/cardio_raw_data.csv"))
```

## Data Exploration Approach

This dataset was obtained from a GitHub repository. The initial exploration included viewing the head of the data to get an overview of the variables, inspecting the data types, and examining the tail of the dataset to identify any potential inconsistencies or anomalies.

We also checked the dimensions of the dataset to ensure that all variables had the same number of observations. To clean the data, we addressed any missing values (NA values) and verified the integrity of categorical variables. This was done by summarizing the data, using `count(n)` for categorical variables to detect any unexpected entries, and calculating the mean for numerical variables to assess and handle missing data appropriately.

```{r}

# Viewing the first rows 
head(Cardiac_raw_data) #Looks like everything is good at the first 7 rows
glimpse(Cardiac_raw_data) # Age: int, gender: chr, height: int, weight: dbl, Sys: int, dias: int, Cholesterol: chr, Glucose: chr, smoke: chr, alcohol: chr, physical: chr, disease: chr, bmi: dbl
tail(Cardiac_raw_data)

# Checking dimensions 
dim(Cardiac_raw_data) #68,675 rows, 13 colums 

# looking at ages 
Cardiac_raw_data |> 
  count(age) 

# looking at height
Cardiac_raw_data |> 
  count(height) 

# looking at weight
Cardiac_raw_data |> 
  count(weight) 

# Sumarise to see if there is any NA
Cardiac_raw_data |> 
  summarise(mean_age = mean(age),
            mean_height = mean(height),
            mean_weight = mean(weight),
            mean_systolic = mean(systolic),
            mean_diastolic = mean(diastolic),
            mean_bmi = mean(bmi)) #No NAs found 

#How many times each chr variable appers, and if there is any miswriting 
Cardiac_raw_data |>
  group_by(gender) |> # female 44,737 male 23,938  = 68,675 
  summarise(count = n())

Cardiac_raw_data |>
  group_by(cholesterol) |> #Extermely High 7,864 High 9,296 Normal 51,515 = 68,675 
  summarise(count = n())
  

Cardiac_raw_data |>
  group_by(glucose) |> # Extremely High 5,218 High 5,059 Normal 58,398 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(smoke) |> #NO 62,635 Yes 6,040 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(alcohol_intake) |> # No 65,007 Yes 3,668  = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(physical_activity) |> #No 13,509 Yes 55,166 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(cv_disease) |> # No 34,724 Yes 33,951 = 68,675
  summarise(count = n())
```

## Initial Data Tidying

We began by creating vectors for the categorical variables. In this step, we defined binary vectors for "Yes" and "No" responses, gender ("Female" and "Male"), and cholesterol and glucose levels, categorized as "Extremely High," "High," and "Normal."

After creating these vectors, we saved the tidied dataset into a new object in preparation for exporting it as a csv file. Within this object, we mutated the categorical variables into factors, renamed them for clarity, and reordered the columns to improve the dataset's presentation and readability.

```{r}
#Creating vectors that would be converted into factors later 
Vector_yes_no <- c("Yes", "No") #This vector would be use into the Yes No answers for Smoking, alcohol, physical activity and cardiovascular disease 

Vector_levels_glu_and_cho <-  c("Extremely High", "High", "Normal") #This vector would be use for the answers in Cholesterol and glucose levels

Vector_gender <- c("Female", "Male") #this vector would be use into the gender answers 

# Creating the new tidied data set 
Tidied_cardiac_data <- Cardiac_raw_data |> 
  #Mutating from chr to factor
  mutate(gender_fct = factor(gender, levels = Vector_gender),
         cholesterol_fct = factor(cholesterol, levels = Vector_levels_glu_and_cho),
         glucose_fct = factor(glucose, levels= Vector_levels_glu_and_cho),
         smoke_fct = factor(smoke, levels = Vector_yes_no),
         alcohol_fct = factor(alcohol_intake, levels = Vector_yes_no),
         physical_activity_fct = factor(physical_activity, levels = Vector_yes_no),
         cv_disease_fct = factor(cv_disease, levels = Vector_yes_no)) |> 
  #Selecting the new column factors and numeric variables
  select(age, 
         gender_fct, 
         height, 
         weight, 
         systolic, 
         diastolic, 
         cholesterol_fct, 
         glucose_fct, 
         smoke_fct, 
         alcohol_fct, 
         physical_activity_fct, 
         cv_disease_fct, 
         bmi) |> 
  #Renaming the colums 
  rename(gender = gender_fct,
         cholesterol = cholesterol_fct,
         glucose = glucose_fct,
         smoke = smoke_fct,
         alcohol_intake = alcohol_fct,
         physical_act= physical_activity_fct,
         Cardiovascular_disease = cv_disease_fct) |> 
  #Relocating bmi after diastolic 
  relocate(bmi, .after = diastolic)

#Saving results 
write_csv(Tidied_cardiac_data, here::here("data/processed/tidied_Cardiac_data.csv"))
```

### Data visualization.

First of All, we will read the tidied dataset that we created in the previous section.

```{r}
Tidied_cardiac_data <- read.csv(here::here("data/processed/tidied_Cardiac_data.csv"))
```

In this section, we present visualizations of the dataset focused on extracting relevant insights. Specifically, we aim to explore whether there is a relationship between high blood pressure (defined as values above 130/80 mmHg), cholesterol levels, and the presence of cardiovascular disease (CVD).

Additionally, we seek to answer the question of which age groups have more cardiovascular disease, based on the available data.

Finally, we examine the prevalence of cardiovascular disease in relation to lifestyle factors, such as alcohol consumption, smoking, and physical activity.

#### Figure 1.

**Hypertension vs Cholesterol levels and Cardiovascular diseases.**

```{r}
#| label: fig-1
#| fig-cap: Hypertension vs Cholesterol levels and Cardiovascular diseases


High_Pressure_Cholesterol_CV_disease<- Tidied_cardiac_data |> 
  select(systolic, diastolic, cholesterol, Cardiovascular_disease) |> 
  filter(systolic >= 130 & diastolic >= 80) |> 
  mutate(High_Pressure = (systolic + diastolic) / 2)


ggplot(data = High_Pressure_Cholesterol_CV_disease, 
       mapping = aes(x = cholesterol,
                     y = High_Pressure, 
                     fill= Cardiovascular_disease)) + 
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(breaks = seq(100, 145, 5), limits = c(100, 145)) +
  scale_fill_brewer(type = "qual", palette = 5) +
  labs(title = "Hypertension and Cholesterol and Cardiovascular diseases",
       subtitle = "Prevalence of cardiovascular disease in individuals with levels of cholesterol and blood pressure",
       x = "Cholesterol levels",
       y = "High blood pressure (mean mmHg)",
       fill = "Cardiovascular diseases",
       caption = "Data: Tidied Cardiac dataset."
  ) +
  theme_minimal(base_size = 11) 

```

@fig-1 presents a visualization comparing groups of individuals with extremely high, high, and normal cholesterol levels, distinguishing between those with and without cardiovascular disease (CVD). The dataset has been filtered to include only individuals with hypertension, defined as having an average systolic blood pressure above 130 mmHg and diastolic blood pressure above 80 mmHg.

The figure uses a boxplot to display the distribution of mean blood pressure across the different cholesterol level groups, highlighting the average values, identifying outliers, and illustrating the prevalence of CVD within each group.

From the chart, we observe that individuals with high and normal cholesterol levels tend to have higher blood pressure if they also have cardiovascular disease, compared to those without CVD. Notably, these groups also show a higher average blood pressure, around 115 mmHg, among individuals with CVD, while those without CVD have a lower mean of approximately 110 mmHg.

In contrast, for the group with extremely high cholesterol levels, there is no significant difference in mean blood pressure between individuals with and without CVD; both subgroups show an average close to 115 mmHg.

#### Figure 2.

**Age distribution and Cardiovascular disease.**

```{r}
#| label: fig-2
#| fig-cap: Age distribution and Cardiovascular disease.

cv_disease_age <- Tidied_cardiac_data |> 
  select(age, Cardiovascular_disease) 
  
ggplot(data = cv_disease_age,
       mapping = aes(x = age,
                     fill = Cardiovascular_disease)) +
  geom_histogram(col = "grey30",
                 breaks = seq(30, 70, 2)) +
  scale_fill_brewer(type = "qual", palette = 5) +
  scale_y_continuous(breaks = seq(0, 7500, 1000), limits = c(0, 7500)) +
  labs(title = "Age and Cardiovascular diseases",
       subtitle = "Prevalence of Cardiovascular Disease Across Age Groups",
       x = "Age groups (2 years)",
       y = "Number of people",
       fill = "Cardiovascular diseases",
       caption = "Data: Tidied Cardiac dataset."
  ) +
  theme_minimal(base_size = 11) 

```

@fig-2 presents a histogram illustrating the age distribution of the population, grouped in two-year intervals, starting from 38 years old and ending at 66 years old. The chart also differentiates between individuals with and without cardiovascular disease (CVD): light green bars represent people without CVD, while light orange bars represent those with CVD.

The distribution of both groups follows a pattern similar to a Gaussian bell curve, with a noticeable peak in the 56–58 age range. This visualization provides key insights: overall, we can observe that younger individuals are less likely to have CVD, whereas the prevalence of CVD increases with age, reaching almost 80% in the 65–66 age group.

#### Figure 3.

**Cardiac disease with the intake of alcohol, smoking and physical activity.**

```{r}
#| label: fig-3
#| fig-cap: Cardiac disease with the intake of alcohol, smoking and physical activity.


bad_habits <- Tidied_cardiac_data |> 
  select(alcohol_intake, smoke, physical_act, Cardiovascular_disease) |> 
  group_by(alcohol_intake, smoke, physical_act, Cardiovascular_disease) |>
  rename(alcohol = alcohol_intake,
         smoking = smoke, 
         Exercising = physical_act) |> 
  pivot_longer(c("alcohol", "smoking", "Exercising"), names_to = "Habit", values_to = "Y_o_N")



ggplot(data = bad_habits,
       mapping = aes(x = Cardiovascular_disease, 
                     fill = Y_o_N)) +
  geom_bar() +
  facet_wrap(~Habit) +
  scale_size(range = c(2, 12)) +
  scale_y_continuous(breaks = seq(0, 40000, 10000), limits = c(0, 40000)) +
  scale_fill_brewer(type = "qual", palette = 5) +
  labs(title = "Cardiovascular disease regarding unhealthy lifestyle habits",
       subtitle = "presence of Cardiovascular across Alcohol, smoking and physical activity",
       x = "Presence of Cardiovascular diseases",
       y = "Number of people",
       fill = "Prevalence of
Unhealthy Lifestyle
Habits",
       caption = "Data: Tidied Cardiac dataset."
  ) +
  theme_minimal(base_size = 11) 

```

@fig-3 is a grouped bar chart displaying the frequency of individuals engaging in unhealthy habits such as alcohol consumption, lack of physical activity, and smoking. The x-axis represents the presence or absence of cardiovascular disease (CVD), while the bar colors indicate whether or not the individual engages in the activity: light green for *No* and light orange for *Yes*.

Regarding alcohol consumption, the data shows that very few individuals engage in this habit, regardless of whether they have CVD or not. In terms of physical activity, both groups, those with and without CVD, show a high prevalence of regular exercise. Finally, in the case of smoking, only a small portion of individuals in either group report smoking, though this number is slightly higher than that for alcohol consumption.

#### Table 1.

**Statistics.**

```{r}
#| label: tbl-1
#| tbl-cap: Statitics for Age groups and averages for age, systolic, diastolic and bmi 


Tidied_cardiac_data |>  
  filter(age >= 54 & systolic >= 130 & diastolic >= 80 & smoke == "Yes" & alcohol_intake == "Yes" & physical_act == "No" & Cardiovascular_disease == "Yes") |> 
  group_by(age) |>  
  summarise(count = n(),
            mean_age = mean(age),
            mean_systolic = mean(systolic),
            sd_systolic = sd(systolic),
            mean_diastolic = mean(diastolic),
            sd_diastolic = sd(diastolic),
            mean_bmi = mean(bmi),
            sd_bmi = sd(bmi)
            ) |> 
  ungroup() |> 
  gt() |> 
  fmt_number(columns = c(mean_systolic, mean_diastolic),
             decimals = 1)
```

@tbl-1 presents the distribution and averages of key health indicators for individuals aged 54 and above, filtered by specific risk factors. These include: systolic blood pressure over 130 mmHg, diastolic blood pressure over 80 mmHg, smoking status, alcohol consumption, lack of physical activity, presence of cardiovascular disease (CVD), and BMI.

The table displays the mean values per age group, along with the corresponding standard deviations for each variable, providing insight into both central tendencies and data variability among older adults with elevated cardiovascular risk.

# Results

@fig-1 shows the prevalence of cardiovascular diseases among individuals with a mean blood pressure above 110 mmHg and normal or high cholesterol levels. This pattern may be explained by reduced arterial and venous lumen, limiting blood flow, as discussed in @felix2013cholesterol.

However, for individuals with extremely high cholesterol levels, there is no significant difference in the presence or absence of cardiovascular disease. Interestingly, the most uncommon cases appear in the high cholesterol group, particularly among those with cardiovascular disease, driven by one outlier with a mean blood pressure around 145 mmHg.

@fig-2 shows the distribution of ages and the prevalence of cardiovascular diseases (CVDs) across different age ranges. In this case, the data reveals a clear trend: as age increases, the prevalence of CVDs also rises, with a more noticeable increase starting from age 56 and above.

In fact, the older the individuals, the more common cardiovascular diseases become, indicating a strong correlation between aging and the risk of developing CVD.

@fig-3 shows a triple-variable correlation graph, illustrating the frequency and distribution of cardiovascular disease (CVD) in relation to unhealthy habits such as alcohol consumption, smoking, and lack of physical activity.

Interestingly, the data shows that individuals who smoke or consume alcohol appear to have a lower prevalence of CVD. However, this trend is likely due to the small number of individuals in the dataset who report engaging in these habits, which limits the strength of this observation.

In contrast, when it comes to physical activity, there is a clearer pattern: individuals who exercise regularly tend to have a lower prevalence of CVD, while those who do not engage in physical activity show a higher incidence of cardiovascular conditions.

in @tbl-1 presents the distribution and average values of age, systolic and diastolic blood pressure, and BMI. The data shown includes only participants who meet all of the following criteria: high blood pressure (above 130/80 mmHg), are active smokers, consume alcohol, lack physical activity, and have been diagnosed with cardiovascular disease (CVD).

From the table, we can observe that the most common age among this group is 62 years. Notably, none of the individuals have a systolic pressure below 140 mmHg or a diastolic pressure below 83 mmHg, confirming the presence of consistently high blood pressure.

In terms of BMI, all participants have values above 27, which falls into the overweight category. Some individuals even reach BMI values consistent with obesity, highlighting an elevated risk profile across multiple health factors [@khanna2022body].

# Conclusions

## Summary of Findings

-   Individuals with normal and high cholesterol levels and mean blood pressure over 110 mmHg show a higher prevalence of CVD

-   CVD prevalence increases with age, showing a sharp rise from age 56 and above.

-   Due to the small number of individuals reporting alcohol consumption or smoking, their relation to CVD in this dataset appears weak or inconclusive.

-   BMI values were above 27 in all cases, indicating overweight or obesity, which further increases CVD risk.

## Questions and Next Steps

Some of the next steps for further analysis include exploring the relationship between height and weight and their impact on the prevalence of cardiovascular diseases (CVDs), with a particular focus on whether taller or heavier individuals are more susceptible. Additionally, it would be valuable to investigate the role of glucose levels and assess the potential influence of diabetes as a contributing factor to CVD. Another important direction would be to examine the most common unhealthy behaviors by gender, identifying whether patterns such as smoking, alcohol consumption, or lack of physical activity differ between males and females, and how these differences may affect cardiovascular risk.

## Bibliographies
