---
title: "Data exploration for Heart diseases and common habits"
author: 
  - name: "Jose Daniel Trujillo Sandoval"
    orcid: "0009-0008-2978-2604"
    email: "jotr4809@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
editor: visual
---

# Introduction

This project would focus on assessing habits related to heart diseases. The data set contains information of random patients, Systolic and Diastolic heart rates, Body mass index, height and weight, and common habits like drinking and smoking

# Methods

## Reading the Data

Loading Packages

```{r}
# Load Packages
library(tidyverse)
library(here) 
# Read Data 
Cardiac_raw_data <- read.csv(here::here("data/raw/cardio_raw_data.csv"))
```

## Data Exploration Approach

```{r}

# Viewing the first rows 
head(Cardiac_raw_data) #Looks like everything is good at the first 7 rows
glimpse(Cardiac_raw_data) # Age: int, gender: chr, height: int, weight: dbl, Sys: int, dias: int, Cholesterol: chr, Glucose: chr, smoke: chr, alcohol: chr, physical: chr, disease: chr, bmi: dbl

# Checking dimensions 
dim(Cardiac_raw_data) #68,675 rows, 13 colums 

# looking at ages 
Cardiac_raw_data |> 
  count(age) 

# looking at height
Cardiac_raw_data |> 
  count(height) 

# looking at weight
Cardiac_raw_data |> 
  count(weight) 

# Sumarise to see if there is any NA
Cardiac_raw_data |> 
  summarise(mean_age = mean(age),
            mean_height = mean(height),
            mean_weight = mean(weight),
            mean_systolic = mean(systolic),
            mean_diastolic = mean(diastolic),
            mean_bmi = mean(bmi)) #No NAs found 

#How many times each chr variable appers, and if there is any miswriting 
Cardiac_raw_data |>
  group_by(gender) |> # female 44,737 male 23,938  = 68,675 
  summarise(count = n())

Cardiac_raw_data |>
  group_by(cholesterol) |> #Extermely High 7,864 High 9,296 Normal 51,515 = 68,675 
  summarise(count = n())
  

Cardiac_raw_data |>
  group_by(glucose) |> # Extremely High 5,218 High 5,059 Normal 58,398 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(smoke) |> #NO 62,635 Yes 6,040 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(alcohol_intake) |> # No 65,007 Yes 3,668  = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(physical_activity) |> #No 13,509 Yes 55,166 = 68,675
  summarise(count = n())

Cardiac_raw_data |>
  group_by(cv_disease) |> # No 34,724 Yes 33,951 = 68,675
  summarise(count = n())
```

## Initial Data Tidying

# Results

\[This will be the core of your analysis with specific requirements\]

# Conclusions

## Summary of Findings

## Questions and Next Steps
